A New Metric for the Analysis of Swarms using Potential Fields

### 2018-10-18 paper.tex
Revisions up to end of section IV

The remainder of this paper is structured as follows:
Section~\ref{Section:Related} discusses different metrics that are currently
used to analyse swarms; Section~\ref{Section:SwarmModelling} introduces a basic
swarm model to produce an environment that can be analysed and evaluated using
the two metrics; Section~\ref{Section:StabilityMagnitude} discusses the
mathematical model that produces the swarming effect and how the potential
fields are applied; Section~\ref{Section::Movement1} discusses how the model is
implemented over time to produce movement;
Section~\ref{Section:StabilityNullVector} discusses the effect of the model on
swarm stability and the effects models have on swarm movements; 
% Section~\ref{Section:Jitter} discusses the effect the mathematical model can
% have upon the structure of a swarm in terms of potentially wasted energy. 
Section~\ref{Section:DistanceDynamics} discusses a distance-based metric for
swarm analysis; Section~\ref{Section:MagnitudeDynamics} introduces the new
\textit{cohesion/repulsion-based} metric, showing how it is calculated and
applied; Section~\ref{Section:ApplicationFloodFilling} compares the two metrics
using an area-filling scenario;
Section~\ref{Section:MagnitudeDistanceComparison} discusses the results of
experiments, and, finally, Section~\ref{Section:FutureWork} discusses
additional work that needs to be carried out to exploit the additional
information that the new metric provides.

@@

%% \section{Metric principles\label{section:MagnitudeDynamics}}
%% This section discusses the theory behind swarm movement and the effects that can be measured using both the metrics described above.
\section{Swarm modelling}\label{Section:SwarmModelling}
Currently, much swarm research uses field effects as the method of modelling
inter-agent interactions~\cite{BAF:06, BAFVM:06, BM:09, APZDAMC:09, GP:02,
GP:04, GP:04a, GP:05, GP:11, MYP:09}. The models usually use two field effects
to implement the swarming characteristic. These effects are \textit{cohesion},
to draw agents closer, and \textit{repulsion} to prevent agents colliding.
\textit{Potential fields} are ranges around an agent that determine the effect
other agents have upon its movement~(Fig.~\ref{methods:FieldEffects}). It is
usual for the cohesion field to have a radius $C_b$ which is larger than the
repulsion radius $R_b$. When an agent $b'$ moves into the \textit{cohesion
field} of an agent $b$, then $b'$ is said to be a neighbour of $b$, and $b$ is
subject to cohesion. When an agent $b'$ moves into the \textit{repulsion field}
of $b$, then $b$ is subject to repulsion. When the repulsion magnitude exceeds
the cohesion magnitude, the agent $b$ will have a tendency to move away from
$b'$, i.e. to be repelled. Obstacles can be introduced into the model. When an
agent $b$ moves too close to an obstacle, i.e. within the obstacle repulsion
range $O_b$, it has a \textit{repulsion vector} applied, giving it a tendency
to move away from the obstacle.

@@

\section{Inter-agent vector magnitude effect on internal movement}\label{Section:StabilityMagnitude}
Figure~\ref{methods:Stability5} shows the cohesion and repulsion vector
contributions of neighbour $b'$ to $v_c(b)$ and $v_r(b)$, as given in
Equations~\ref{eq:FlyToCentre1} and \ref{eq:Repulsion1}. Notice that the vectors
are along the line of separation $bb'$. \\

@@

\section{Internal movement and the null vector}\label{Section:StabilityNullVector}

@@

If we consider the equilibrium state (Fig.~\ref{methods:Stability4}) the
resultant $v(b)=(0,0)$. A null vector cannot be normalised to produce a
directional vector ($\hat{v} = \frac{v}{|v|}$ if $v\neq0$; $0$ if $v=0$) The
effect is that the agent will remain stationary. If all agent pairs are in this
condition the swarm will stop moving~(Fig.~\ref{methods:StabilityNullVector}).

@@

Due to the movement of agents being based only on neighbour interactions this
situation is very rare. The residual motion that persists in a swarm is the
background `noise' or `jitter' that an algorithm creates.

If a swarm is goal-based, the addition of a \textit{directional vector} will
prevent all agents simultaneously producing null vectors as shown
in~Fig.~\ref{concave:VoidPerimeter2} and
Equation~\ref{eq:InterAgentMovementDirectional1}. Here, $k_dv_{d}(b)$ is the
calculated directional vector to be applied to an agent.

\begin{equation}\label{eq:InterAgentMovementDirectional1}
v(b) = k_cv_{c}(b) + k_rv_{r}(b) + k_dv_{d}(b)
\end{equation}

@@

\section{Modelling movement}\label{Section::Movement1}
Each agent within a swarm calculates its \textit{movement-direction vector}
based on its \textit{interaction} and \textit{destination} vectors. The
\textit{movement vector}~($b_{pos}$) is calculated using the unit
\textit{movement-direction vector}
(Equation~\ref{eq:InterAgentMovementDirectional1}) multiplied by the time
elapsed~($t$) in the system and the speed characteristic of the agent~($s_b$).

@@

The increment in the location of agent $b$ over time interval $t$ is shown in
Equation~\ref{eq:AgentMovement} where $s_b$ is the speed of agent~$b$. Note
that \string^ is the equivalent of $\hat{v} = \frac{v}{|v|}$ the normalised
vector.

\section{Swarm movement analysis\label{metric:MagnitudeDynamics2}}
The repulsion and cohesion vectors are generated for an agent through the
interaction of their field effects~(Fig.~\ref{methods:FieldEffects}). There are
a limited number of interactions that can occur; These are illustrated in
Fig.~\ref{methods:Stability1}, \ref{methods:Stability2},
\ref{methods:Stability3} and \ref{methods:Stability4}.

Following repeated experiments in the initial deployment of a swarm it was
found that using the same parameters for a swarm's environmental parameters and
the cohesion and repulsion ranges that the results where always very similar.
The results shown in this paper are therefore an example set typical of the
results obtained. The data extracts shown in Tables \ref{tab:SampleReplusion0},
\ref{tab:SampleReplusionPositive}, \ref{tab:SampleCohesionPositive} and
\ref{tab:SampleEquilibrium} are the simulation results that are produced by the
parameters listed in Table~\ref{tab:MetricPhysics1}. The simulation consists of
200 agents over a 20 second period. The cohesion of an agent pair is shown as
$k_cv_c$ and the repulsion as $k_rv_r$.

@@

Figure~\ref{methods:Stability1} shows two agents within each other's cohesion
fields but sufficiently distant to be outside of the repulsion fields. In this
case $|k_cv_c| > 0$ and $|k_rv_r| = 0$: the result is the agent's resultant
magnitude $v(b)$ causes the agent to move towards its neighbour $b'$. Likewise
the neighbour's resultant magnitude $v(b')$ will cause it to move towards $b$.
Table~\ref{tab:SampleReplusion0} shows the repulsion magnitude with a value of
0. The only influence on the agent pair are cohesive vectors.

@@

Figure~\ref{methods:Stability4} shows two agents close together with
$|k_cv_c|~=~|k_rv_r|$ the resultant vector is a \textit{null vector} and the
agents have no overall influence upon each other due to the magnitude of the
resultant vector being zero. Table~\ref{tab:SampleEquilibrium} is an extract of
data from the simulator. The data shows near equilibrium but due to the dynamic
nature of a swarm system no agents meet the condition fully.

@@

Section~\ref{Section:StabilityMagnitude}~discussed calculation of resultant
cohesion~(Equation~\ref{eq:FlyToCentre1}) and resultant
repulsion~(Equation~\ref{eq:Repulsion1}) vectors for each agent deriving a
combined \textit{cohesion/repulsion vector} using as shown in
Equation~\ref{eq:InterAgentMovement2}.

>>>>

The remainder of this paper is structured as follows:
Section~\ref{Section:Related} discusses different metrics that are currently
used to analyse swarms; Section~\ref{Section:SwarmModelling} introduces a 
model of swarm behaviour, using potential fields, that can be analysed and evaluated using
the two metrics; Section~\ref{Section:SwarmDynamics} shows examples of the typical
dynamic behaviour of swarms under the potential field model;
% Section~\ref{Section:Jitter} discusses the effect the mathematical model can
% have upon the structure of a swarm in terms of potentially wasted energy. 
Section~\ref{Section:DistanceDynamics} discusses a distance-based metric for
swarm analysis; Section~\ref{Section:MagnitudeDynamics} introduces the new
\textit{cohesion/repulsion-based} metric, showing how it is calculated and
applied; Section~\ref{Section:ApplicationFloodFilling} compares the two metrics
using an area-filling scenario;
Section~\ref{Section:MagnitudeDistanceComparison} discusses the results of
experiments, and, finally, Section~\ref{Section:FutureWork} discusses
additional work that needs to be carried out to exploit the additional
information that the new metric provides.

@@

%% \section{Metric principles\label{section:MagnitudeDynamics}}
%% This section discusses the theory behind swarm movement and the effects that can be measured using both the metrics described above.
\section{Swarm modelling with potential fields}\label{Section:SwarmModelling}
Currently, much swarm research uses field effects as the method of modelling
inter-agent interactions~\cite{BAF:06, BAFVM:06, BM:09, APZDAMC:09, GP:02,
GP:04, GP:04a, GP:05, GP:11, MYP:09}. The models usually use two field effects
to implement the swarming characteristic. These effects are \textit{cohesion},
to draw agents closer, and \textit{repulsion} to prevent agents colliding.
\textit{Potential fields} are ranges around an agent that determine the effect
other agents have upon its movement~(Fig.~\ref{methods:FieldEffects}). It is
usual for the cohesion field to have a radius $C_b$ which is larger than the
repulsion radius $R_b$. When an agent $b'$ moves into the \textit{cohesion
field} of an agent $b$, then $b'$ is said to be a neighbour of $b$, and $b$ is
subject to cohesion. When an agent $b'$ moves into the \textit{repulsion field}
of $b$, then $b$ is subject to repulsion. When the repulsion magnitude exceeds
the cohesion magnitude, the agent $b$ will have a tendency to move away from
$b'$, i.e. to be repelled. Obstacles can be introduced into the model. When an
agent $b$ moves too close to an obstacle, i.e. within the obstacle repulsion
range $O_b$, it has a \textit{repulsion vector} applied, giving it a tendency
to move away from the obstacle.

@@

%\section{Inter-agent vector magnitude effect on internal movement}\label{Section:StabilityMagnitude}
Figure~\ref{methods:Stability5} shows the cohesion and repulsion vector
contributions of neighbour $b'$ to $v_c(b)$ and $v_r(b)$, as given in
Equations~\ref{eq:FlyToCentre1} and \ref{eq:Repulsion1}. Notice that the vectors
are along the line of separation $bb'$. \\

@@

%\section{Internal movement and the null vector}\label{Section:StabilityNullVector}

@@

If we consider the equilibrium state, % (Fig.~\ref{methods:Stability4})
the resultant $v(b)=(0,0)$. A null vector cannot be normalised to produce a
directional vector ($\hat{v} = \frac{v}{|v|}$ if $v\neq0$; $0$ if $v=0$) The
effect is that the agent will remain stationary. If all agent pairs are in this
condition the swarm will stop moving~(Fig.~\ref{methods:StabilityNullVector}).

@@

Due to the movement of agents being based only on neighbour interactions, this
situation is very rare. The residual motion that persists in a swarm is the
background `noise' or `jitter' that an algorithm creates.

A swarm can also have a \emph{goal}. This is modelled simply by the addition of 
a \textit{directional vector}, as shown in~Equation~\ref{eq:InterAgentMovementDirectional1}.
\begin{equation}\label{eq:InterAgentMovementDirectional1}
v(b) = k_cv_{c}(b) + k_rv_{r}(b) + k_dv_{d}(b)
\end{equation}
Here, $k_dv_{d}(b)$ is the
calculated directional vector to be applied to an agent.

Goal-based swarms tend to 
prevent all agents simultaneously producing null vectors, as suggested by
Fig.~\ref{concave:VoidPerimeter1}~and~\ref{concave:VoidPerimeter2}.

@@

\section{Swarm dynamics}\label{Section:SwarmDynamics}
%\section{Modelling movement}\label{Section::Movement1}
Each agent within a swarm calculates its \textit{movement-direction vector}
based on its \textit{inter-agent} and \textit{directional} vectors. The
\textit{movement vector}~($b_{pos}$) is calculated using the unit
\textit{movement-direction vector}
(Equation~\ref{eq:InterAgentMovementDirectional1}), multiplied by the time
elapsed~($t$) in the system, and the speed characteristic of the agent~($s_b$).

@@

The change in the location of agent $b$, over time interval $t$, is shown in
Equation~\ref{eq:AgentMovement}, where $s_b$ is the speed of agent~$b$. Note
that \string^ is the equivalent of $\hat{v} = \frac{v}{|v|}$, the normalised
vector.

The repulsion and cohesion vectors are generated for an agent through the
interaction of its field effects~(Fig.~\ref{methods:FieldEffects}). There are
a limited number of interactions that can occur. These are illustrated in
Fig.~\ref{methods:Stability1}, \ref{methods:Stability2},
\ref{methods:Stability3} and \ref{methods:Stability4}.

Following repeated experiments in the initial deployment of a swarm, it was
found that, using the same parameters for a swarm's environment and
cohesion and repulsion ranges, always produced very similar behaviour.
The results shown in this section are an example set typical of the
results obtained. The data extracts shown in Tables \ref{tab:SampleReplusion0},
\ref{tab:SampleReplusionPositive}, \ref{tab:SampleCohesionPositive} and
\ref{tab:SampleEquilibrium} are the simulation results that are produced by the
parameters listed in Table~\ref{tab:MetricPhysics1}. The simulation consists of
200 agents over a 20 second period. The cohesion of an agent pair is shown as
$k_cv_c$ and the repulsion as $k_rv_r$.

@@

Figure~\ref{methods:Stability1} shows two agents within each other's cohesion
fields but sufficiently distant to be outside of the repulsion fields. In this
case $|k_cv_c| > 0$ and $|k_rv_r| = 0$: the result is that the agent's resultant
magnitude $v(b)$ causes the agent to move towards its neighbour $b'$. Likewise
the neighbour's resultant magnitude $v(b')$ will cause it to move towards $b$.
Table~\ref{tab:SampleReplusion0} shows the repulsion magnitude with a value of
0. The only influence on the agent pair are cohesive vectors.

@@

Figure~\ref{methods:Stability4} shows two agents close together with
$|k_cv_c|~=~|k_rv_r|$. The resultant vector is a \textit{null vector} and the
agents have no overall influence upon each other, since the magnitude of the
resultant vector is zero. Table~\ref{tab:SampleEquilibrium} is an extract of
data from the simulator. The data shows near equilibrium, but due to the dynamic
nature of a swarm system, no agents meet the condition fully.

@@

Section~\ref{Section:SwarmModelling}~discussed calculation of resultant
cohesion~(Equation~\ref{eq:FlyToCentre1}) and resultant
repulsion~(Equation~\ref{eq:Repulsion1}) vectors for each agent deriving a
combined \textit{cohesion/repulsion vector} using as shown in
Equation~\ref{eq:InterAgentMovement2}.
